I am faced with a basic problem of presentation. What I am really 
concerned about is the composition of large programs, the text of 
which may be, say, of the same size as the whole text of this 
booklet. Also I have to include examples to illustrate the various
techniques. For practical reasons, the demonstration programs 
must be small, many times smaller than the ``life-size program''
I have in mind. My basic problem is that precisely this difference 
in scale is one of the major sources of our difficulties in 
programming!

It would be very nice if I could illustrate the various techniques 
with small demonstration programs and could conclude with ``... 
and when faced with a program a thousand times as large, you 
compose it in the same way.'' This common educational device, 
however, would be self-defeating as one of my central themes will 
be that any two things that differ in some respect by a factor 
of already a hundred or more, are utterly incomparable.

History has shown that this truth is very hard to believe. 
Apparently we are too much trained to disregard differences in 
scale, to treat them as ``gradual differences that are not 
essential''. We tell ourselves that what we can do once, we can 
also do twice and by induction we fool ourselves into believing 
that we can do it as many times as needed, but this is just not 
true! A factor of a thousand is already far beyond our powers of
imagination!

Let me give you two examples to rub this in. A one-year old child
will crawl on all fours with a speed of, say, one mile per hour. 
But a speed of a thousand miles per hour is that of a super sonic 
jet. Considered as objects with moving ability the child and the 
jet are incomparable, for whatever one can do the other cannot and 
vice versa. Also: one can close one's eyes and imagine how it feels 
to be standing in an open place, a prairie or a sea shore, while 
far away a big, reinless horse is approaching at a gallop, one can
``see'' it approaching and passing. To do the same with a phalanx 
of a thousand of these big beasts is mentally impossible: your 
heart would miss a number of beats by pure panic, if you could!

To complicate matters still further, problems of size do not only 
cause me problems of presentation, but they lie at the heart of the 
subject: widespread undersetimation of the specific difficulties 
of size seems one of the major underlying causes of the current 
software failure. To all this I can see only one answer, viz. to 
treat problems of size as explicitly as possible. Hence the title 
of this section.

To start with, we have the ``size'' of the computation, i.e. the 
amount of information and the number of operations involved in it. 
It is essential that this size is large, for if it were really 
small, it would be easier not to use the computer at all and to do 
it by hand. The automatic computer owes it right to exist, its 
usefulness, precisely to its ability to perform large computations
where we humans cannot. We want the computer to do what we could 
never do ourselves and the power of present-day machinery is such
that even small computations are by their very size already far 
beyond the powers of our unaided imagination.

Yet we must organize the computations in such a way that our limited 
powers are sufficient to guarantee that the computation will 
establish the desired effect. This organizing includes the
composition of the program and here we are faced with the next 
problem of size, viz. the length of the program text, and we should 
give this problem also explicit recognition. We should remain aware 
of the fact that the extent to which we can read or write a text is 
very much dependent on its size. In my country the entries in the 
telephone directory are grouped by town or village and within each 
such group the subscribers are listed by name in alphabetical
order. I myself live in a small village and given a telephone number 
I have only to scan a few columns to find out to whom the telephone 
number belongs, but to do the same in a large city would be a major 
data processing task!

It is in the same mood that I should like to draw the reader's 
attention to the fact that ``clarity'' has pronounced quantitative 
aspects, a fact many mathematicians, curiously enough, seem to be 
unaware of. A theorem stating the validity of a conclusion when ten 
pages full of conditions are satisfied is hardly a convenient tool, 
as all conditions have to be verified whenever the theorem is 
appealed to. In Euclidean geometry, Pythagoras' Theorem holds for any 
three points A, B and C such that through A and C a straight line can 
be drawn orthogonal to a straight line through B and C. How many 
mathematicians appreciate that the theorem remains applicable when 
some or all of the points A, B and C coincide? Yet this seems largely 
responsible for the convenience with which Pythagoras Theorem can be
used.

Summarizing: as a slow-witted human being I have a very small head 
and I had better learn to live with it and to respect my limitations 
and give them full credit, rather than to try to ignore them, for the 
latter vain effort will be punished by failure.